---
title: "Lab 1"
author: "Mark Tello-Rincon"
format: html
editor: visual
embed-resources: true
theme: cerulean
fig-width: 7
fig-height: 7
---

```{r}
library(datasauRus)
```

## Question 1

```{r}
?datasaurus_dozen
```

There are 1846 rows and 3 columns.

The variables included in the data frame are:

-   dataset
-   x
-   y

## Question 2

#### Subsetting

```{r}
dino_data <- datasaurus_dozen[datasaurus_dozen$dataset == 'dino', ]
```

#### Plotting

```{r}
plot(dino_data$x, dino_data$y)
# ggplot(data = dino_data, mapping = aes(x = x, y = y)) +
#   geom_point()
```

#### Correlation Coefficient

```{r}
cor(dino_data$x, dino_data$y)
# dino_data |>
#   summarize(r = cor(x, y))
```

## Question 3

#### Subsetting

```{r}
star_data <- datasaurus_dozen[datasaurus_dozen$dataset == 'star', ]
```

#### Plotting

```{r}
plot(star_data$x, star_data$y)
# ggplot(data = star_data, mapping = aes(x = x, y = y)) +
#   geom_point()
```

#### Correlation Coefficient

```{r}
cor(star_data$x, star_data$y)
# star_data |>
#   summarize(r = cor(x, y))
```

The `dino_data` correlation coefficient is -0.06447185, while the `star_data` correlation coefficient is -0.0629611.

## Question 4

#### Subsetting

```{r}
circle_data <- datasaurus_dozen[datasaurus_dozen$dataset == 'circle', ]
```

#### Plotting

```{r}
plot(circle_data$x, circle_data$y)
# ggplot(data = circle_data, mapping = aes(x = x, y = y)) +
#   geom_point()
```

#### Correlation Coefficient

```{r}
cor(circle_data$x, circle_data$y)
# circle_data |>
#   summarize(r = cor(x, y))
```

The `dino_data` correlation coefficient is -0.06447185, while the `circle_data` correlation coefficient is -0.06834336.

## Question 5

```{r}
#| fig-height: 15
#| fig-width: 15
layout(matrix(1:16, nrow=4, ncol=4))
for(name in unique(datasaurus_dozen$dataset)){
  subset <- datasaurus_dozen[datasaurus_dozen$dataset == name, ]
  plot(subset$x, subset$y, main = name)
}
layout(1)
```

```{r}
# ggplot(datasaurus_dozen, aes(x = x, y = y, color = dataset))+
#   geom_point()+
#   facet_wrap(~ dataset, ncol = 3) +
#   theme(legend.position = "none")
```

## Question 6

```{r}
sapply(unique(datasaurus_dozen$dataset), function(name){
    subset <- datasaurus_dozen[datasaurus_dozen$dataset == name, ]
    return(cor(subset$x, subset$y))
})
```

```{r}
# datasaurus_dozen |>
#   group_by(dataset) |>
#   summarize(r = cor(x, y))
```
